---
title: "20250219 Tables and Figures"
output:
  html_document: default
  word_document: default
date: "2025-02-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(include = FALSE)
knitr::opts_chunk$set(results = FALSE)
```

## Setup

```{r load packages}
library(table1)
library(Hmisc)
library(tidyverse)
library(gdata)
library(xtable)
```

```{r load data}
today <- Sys.Date()
```

```{r load data}
load(paste0("./infile/", today, "_cognition_data_final_out.RData"))
```

```{r load results}
load(file = paste0("./output/", today, "_all.RData"))
load(file = paste0("./output/", today, "_male.RData"))
load(file = paste0("./output/", today, "_female.RData"))
load(file = paste0("./output/", today, "_hispanic.RData"))
load(file = paste0("./output/", today, "_not_hispanic.RData"))
load(file = paste0("./output/", today, "_white.RData"))
load(file = paste0("./output/", today, "_black.RData"))
load(file = paste0("./output/", today, "_other.RData"))
load(file = paste0("./output/", today, "_all_edu_sens.RData"))
load(file = paste0("./output/", today, "_all_nnet.RData"))
load(file = paste0("./output/", today, "_all_param.RData"))
```

## Function definitions

```{r functions, eval=TRUE}
relevel_table1 <- function(data){
  data_releveled <- data %>% 
    mutate(education = ifelse(education == "college", "College or Higher", "Below College"),
           education_p = ifelse(education_p == "one_parent_above_8_yedu",
                                "One or More Parents With >8 Years of Education",
                                "No Parent With >8 Years of Education"),
           RAGENDER = ifelse(RAGENDER == '1.Male', "Male", "Female"),
           RARACEM = as_factor(case_when(
             RARACEM == "1.White/Caucasian" ~ "White",
             RARACEM == "2.Black/African American" ~ "Black",
             RARACEM == "3.Other" ~ "Other")),
           RAHISPAN = as_factor(case_when(
             RAHISPAN == "0.Not Hispanic" ~ "Non-Hispanic",
             RAHISPAN == "1.Hispanic" ~ "Hispanic")),
           FAMFIN = as_factor(case_when(
             FAMFIN == "WELL" ~ "Well",
             FAMFIN == "AVERAGE" ~ "Average",
             FAMFIN %in% c("POOR", "VOLATILE") ~ "Poor/Volatile")),
           FJOB = as_factor(case_when(
             FJOB == "MANAGERIAL" ~ "Managerial",
             FJOB == "SALES" ~ "Sales",
             FJOB == "CLERICAL" ~ "Clerical",
             FJOB == "SERVICE" ~ "Service",
             FJOB == "MANUAL" ~ "Manual",
             FJOB == "ARMED" ~ "Armed")),
           wave = as_factor(wave),
           HACOHORT = as_factor(HACOHORT))
  
  # tidy labels
  label(data_releveled$RAGEY_B) <- "Age"
  label(data_releveled$RAGENDER) <- "Gender"
  label(data_releveled$education) <- "Education"
  label(data_releveled$education_p) <- "Parental Education"
  label(data_releveled$RCOG27) <- "Cognitive Function"
  label(data_releveled$RARACEM) <- "Race"
  label(data_releveled$RAHISPAN) <- "Ethnicity"
  label(data_releveled$RTHLTHCH) <- "Self-Rated Health During Childhood"
  label(data_releveled$FJOB) <- "Fathers Occupation at 16"
  label(data_releveled$FAMFIN) <- "Financial Situation of the Family During Childhood"
  label(data_releveled$HACOHORT) <- "Cohort"
  label(data_releveled$wave) <- "Wave"
  
  return(data_releveled)
}

render_continuous <- function(x, cutoff = .05, ...) {
  strs <- c(
    "",
    "M (95% CI)" = paste0(
      sprintf("%1.1f", mean(x, na.rm = TRUE)),
      " (",
      sprintf("%1.1f", mean_cl_normal(x, na.rm = TRUE)[2]),
      "; ",
      sprintf("%1.1f", mean_cl_normal(x, na.rm = TRUE)[3]),
      ")")
    )
  
  return(strs)
}

pvalue <- function(x, ...) {
  # Construct vectors of data y, and groups (strata) g
  y <- unlist(x)
  g <- factor(rep(1:length(x), times=sapply(x, length)))
  if (is.numeric(y)) {
    # For numeric variables, perform a standard 2-sample t-test
    p <- t.test(y ~ g)$p.value
  } else {
    # For categorical variables, perform a chi-squared test of independence
    p <- chisq.test(table(y, g))$p.value
  }
  # Format the p-value, using an HTML entity for the less-than sign.
  # The initial empty string places the output on the line below the variable label.
  c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}

rd <- function(x) {format(round(x, 3), nsmall = 3)}

irl <- function(x,y,z) {
  paste(rd(x), paste("(", paste(rd(y), rd(z), sep=", "), ")", sep=""))
}

```


## Filtering

```{r filter data, eval=TRUE, results=TRUE}
cognition_data_final_out_filtered <- cognition_data_final_out %>% 
  # restrict age
  filter(RAGEY_B >= 50 & RAGEY_B <= 60) %>% 
  # id indicator
  group_by(HHIDPN) %>% 
  mutate(ind = 1:n()) %>% 
  ungroup() %>% 
  filter(ind == 1)
```

## Table A1 (Imputed, Analytic Data Set)

```{r table a1, eval=TRUE, results = TRUE, include = TRUE}
table_1_data <- relevel_table1(cognition_data_final_out_filtered)
table1(~ RAGEY_B + education  + RAGENDER +
         RARACEM + RAHISPAN + 
         RCOG27 +
         FJOB + FAMFIN + 
         RTHLTHCH +
         wave |
         education_p, data = table_1_data, overall = F,
       render.continuous = render_continuous)
```

## Table A2 Decomposition estimates

This table is adapted from: https://github.com/ang-yu/causal_decomposition_case_study/blob/main/Analysis

```{r table a2, eval=TRUE, results='asis', message=F, echo=FALSE, warning=FALSE, include = TRUE}
tableA2 <- cbind(
  c("Total",
    "Baseline",
    "Prevalence",
    "Effect",
    "Selection",
    "Jackson reduction"),
  # all
  c(irl(result_all$results$point, result_all$results$CI_lower, result_all$results$CI_upper),
    irl(result_all$results_specific$point[13],
        result_all$results_specific$CI_lower[13],
        result_all$results_specific$CI_upper[13])),
  # female
  c(irl(result_female$results$point, result_female$results$CI_lower, result_female$results$CI_upper),
    irl(result_female$results_specific$point[13],
        result_female$results_specific$CI_lower[13],
        result_female$results_specific$CI_upper[13])),
  # male
  c(irl(result_male$results$point, result_male$results$CI_lower, result_male$results$CI_upper),
    irl(result_male$results_specific$point[13],
        result_male$results_specific$CI_lower[13],
        result_male$results_specific$CI_upper[13])),
  # white
  c(irl(result_white$results$point, result_white$results$CI_lower, result_white$results$CI_upper),
    irl(result_white$results_specific$point[13],
        result_white$results_specific$CI_lower[13],
        result_white$results_specific$CI_upper[13])),
  # black
  c(irl(result_black$results$point, result_black$results$CI_lower, result_black$results$CI_upper),
    irl(result_black$results_specific$point[13],
        result_black$results_specific$CI_lower[13],
        result_black$results_specific$CI_upper[13])),
  # other
  c(irl(result_other$results$point, result_other$results$CI_lower, result_other$results$CI_upper),
    irl(result_other$results_specific$point[13],
        result_other$results_specific$CI_lower[13],
        result_other$results_specific$CI_upper[13])),
  # hispanic
  c(irl(result_hispanic$results$point, result_hispanic$results$CI_lower, result_hispanic$results$CI_upper),
    irl(result_hispanic$results_specific$point[13],
        result_hispanic$results_specific$CI_lower[13],
        result_hispanic$results_specific$CI_upper[13])),
  # not hispanic
  c(irl(result_not_hispanic$results$point, result_not_hispanic$results$CI_lower, result_not_hispanic$results$CI_upper),
    irl(result_not_hispanic$results_specific$point[13],
        result_not_hispanic$results_specific$CI_lower[13],
        result_not_hispanic$results_specific$CI_upper[13])))
    
knitr::kable(xtable(tableA2), type = "html", 
             col.names = c("Effect", 
                           "All", 
                           "Female", "Male", 
                           "White", "Black", "Other Race", 
                           "Hispanic", "Not Hispanic"))
```

## Table A3 Group-specific estimates for the decomposition {.tabset}

This table is adapted from: https://github.com/ang-yu/causal_decomposition_case_study/blob/main/Analysis

### All

```{r table a3 all, eval=TRUE, results=TRUE, message=F, echo=FALSE, warning=FALSE, include = TRUE}
result_all$results_specific[3:12,c("point","CI_lower","CI_upper")] %>% 
  as.data.frame() %>% 
  mutate_if(is.numeric, ~round(., 3)) %>% 
  transmute(res = paste0(point, " (", CI_lower, ", ", CI_upper, ")")) %>% 
  rownames_to_column() %>% 
  filter(grepl("ATE", rowname))
```

### Female

```{r table a3 fem, eval=TRUE, results=TRUE, message=F, echo=FALSE, warning=FALSE, include = TRUE}
result_female$results_specific[3:12,c("point","CI_lower","CI_upper")] %>% 
  as.data.frame() %>% 
  mutate_if(is.numeric, ~round(., 3)) %>% 
  transmute(res = paste0(point, " (", CI_lower, ", ", CI_upper, ")")) %>% 
  rownames_to_column() %>% 
  filter(grepl("ATE", rowname))
```

### Male

```{r table a3 mal, eval=TRUE, results=TRUE, message=F, echo=FALSE, warning=FALSE, include = TRUE}
result_male$results_specific[3:12,c("point","CI_lower","CI_upper")] %>% 
  as.data.frame() %>% 
  mutate_if(is.numeric, ~round(., 3)) %>% 
  transmute(res = paste0(point, " (", CI_lower, ", ", CI_upper, ")")) %>% 
  rownames_to_column() %>% 
  filter(grepl("ATE", rowname))
```

### White

```{r table a3 whi, eval=TRUE, results=TRUE, message=F, echo=FALSE, warning=FALSE, include = TRUE}
result_white$results_specific[3:12,c("point","CI_lower","CI_upper")] %>% 
  as.data.frame() %>% 
  mutate_if(is.numeric, ~round(., 3)) %>% 
  transmute(res = paste0(point, " (", CI_lower, ", ", CI_upper, ")")) %>% 
  rownames_to_column() %>% 
  filter(grepl("ATE", rowname))
```

### Black

```{r table a3 bla, eval=TRUE, results=TRUE, message=F, echo=FALSE, warning=FALSE, include = TRUE}
result_black$results_specific[3:12,c("point","CI_lower","CI_upper")] %>% 
  as.data.frame() %>% 
  mutate_if(is.numeric, ~round(., 3)) %>% 
  transmute(res = paste0(point, " (", CI_lower, ", ", CI_upper, ")")) %>% 
  rownames_to_column() %>% 
  filter(grepl("ATE", rowname))
```

### Other Race

```{r table a3 oth, eval=TRUE, results=TRUE, message=F, echo=FALSE, warning=FALSE, include = TRUE}
result_other$results_specific[3:12,c("point","CI_lower","CI_upper")] %>% 
  as.data.frame() %>% 
  mutate_if(is.numeric, ~round(., 3)) %>% 
  transmute(res = paste0(point, " (", CI_lower, ", ", CI_upper, ")")) %>% 
  rownames_to_column() %>% 
  filter(grepl("ATE", rowname))
```

### Hispanic

```{r table a3 his, eval=TRUE, results=TRUE, message=F, echo=FALSE, warning=FALSE, include = TRUE}
result_hispanic$results_specific[3:12,c("point","CI_lower","CI_upper")] %>% 
  as.data.frame() %>% 
  mutate_if(is.numeric, ~round(., 3)) %>% 
  transmute(res = paste0(point, " (", CI_lower, ", ", CI_upper, ")")) %>% 
  rownames_to_column() %>% 
  filter(grepl("ATE", rowname))
```

### Not Hispanic

```{r table a3 nhi, eval=TRUE, results=TRUE, message=F, echo=FALSE, warning=FALSE, include = TRUE}
result_not_hispanic$results_specific[3:12,c("point","CI_lower","CI_upper")] %>% 
  as.data.frame() %>% 
  mutate_if(is.numeric, ~round(., 3)) %>% 
  transmute(res = paste0(point, " (", CI_lower, ", ", CI_upper, ")")) %>% 
  rownames_to_column() %>% 
  filter(grepl("ATE", rowname))
```

## Figure A1 Case numbers and relative share of parental and individual education by race

```{r pct race, results=TRUE, include = TRUE}
cognition_data_final_out_filtered %>% 
  group_by(RARACEM, education_p, education) %>% 
  summarise(n = n()) %>% 
  group_by(RARACEM) %>% 
  mutate(N = sum(n), 
         pct = n/N*100, 
         RARACEM = case_when(
           RARACEM == "1.White/Caucasian" ~ "White",
           RARACEM == "2.Black/African American" ~ "Black",
           RARACEM == "3.Other" ~ "Other"),
         Education = case_when(
           education_p == "one_parent_above_8_yedu" & education == "college" ~ "Higher: Parent(s) & Self",
           education_p == "one_parent_above_8_yedu" & education == "below_college" ~ "Higher: Parent(s) only",
           education_p == "no_parent_above_8_yedu" & education == "college" ~ "Lower: Parent(s) only",
           education_p == "no_parent_above_8_yedu" & education == "below_college" ~ "Lower: Parent(s) & Self",
         )) %>% 
  ggplot(aes(RARACEM, pct, fill = Education, label = paste0("n = ", n))) +
  geom_bar(stat = "identity", lwd = .5, colour = "white", alpha = 1) + 
  theme_minimal() +
  scale_fill_manual(values = c("tomato3", "tomato2", "steelblue2", "steelblue4")) +
  geom_label(position = position_stack(), vjust = 0, alpha = 1, colour = "white") +
  ylab("Percent") + xlab("Self-Reported Race")
```

## Figure A2 Case numbers and relative share of parental and individual education by ethnicity

```{r pct ethn, results=TRUE, include = TRUE}
cognition_data_final_out_filtered %>% 
  group_by(RAHISPAN, education_p, education) %>% 
  summarise(n = n()) %>% 
  group_by(RAHISPAN) %>% 
  mutate(N = sum(n), 
         pct = n/N*100, 
         RAHISPAN = case_when(
           RAHISPAN == "0.Not Hispanic" ~ "Non-Hispanic",
           RAHISPAN == "1.Hispanic" ~ "Hispanic"),
         Education = case_when(
           education_p == "one_parent_above_8_yedu" & education == "college" ~ "Higher: Parent(s) & Self",
           education_p == "one_parent_above_8_yedu" & education == "below_college" ~ "Higher: Parent(s) only",
           education_p == "no_parent_above_8_yedu" & education == "college" ~ "Lower: Parent(s) only",
           education_p == "no_parent_above_8_yedu" & education == "below_college" ~ "Lower: Parent(s) & Self",
         )) %>% 
  ggplot(aes(RAHISPAN, pct, fill = Education, label = paste0("n = ", n))) +
  geom_bar(stat = "identity", lwd = .5, colour = "white", alpha = 1) + 
  theme_minimal() +
  scale_fill_manual(values = c("tomato3", "tomato2", "steelblue2", "steelblue4")) +
  geom_label(position = position_stack(), vjust = 0, alpha = 1, colour = "white") +
  ylab("Percent") + xlab("Self-Reported Ethnicity")
```
